### 2024-04-17 docs/SOSP 2024/hir-mir-opts.tex
Edit overview and schedule sections

\section{\Treebeard{} Multi-Level Compiler}

@@

\Treebeard{} takes a serialized decision tree ensemble as input (
XGBoost JSON, ONNX etc.) and automatically generates an optimized inference function
that can either target CPUs or GPUs. 
Figure \ref{Fig:CompilerStructure} shows the structure of the \Treebeard{} compiler. 
The inference computation is lowered through three intermediate representations
-- high-level IR (HIR), mid-level IR (MIR) and low-level IR (LIR). The LIR is
finally lowered to LLVM and then JIT'ed to the specified target processor. 
\Treebeard{} is built using the open-source \TreebeardOLD{} infrastructure~\cite{Treebeard}.
Since the \TreebeardOLD{} infrastructure was originally designed to target CPUs,
significant extensions were required to support schedule-guided compilation for CPUs and GPUs. 
These parts of the system are shown as shaded boxes in Figure \ref{Fig:CompilerStructure}.
% However, it implements several optimizations on the HIR and MIR that 
% can be leveraged across target processors and we find that some these 
% are beneficial for GPUs as well. 
% This reuse of optimizations is possible 
% because the intermediate representations on which these optimizations are performed 
% are abstract and are designed to be target-independent. We briefly review these 
% optimizations below.

@@

The HIR is lowered to MIR as dictated by the \emph{schedule}.
Optimizations like tree-walk unrolling and tree-walk interleaving
are performed on the MIR. 
% These optimizations are beneficial for GPUs as well.
% and the design of \TreebeardOLD{} allows us to reuse the tree-walk 
% unrolling and tree-walk interleaving optimizations on the MIR for GPUs.
% While building \Treebeard{}, we found that
One surprising thing we found while developing \Treebeard{} was that 
we could use ILP to improve performance on GPUs. 
One of the performance bottlenecks in inference code targeted 
to GPUs was that warps spent significant time being stalled. Since GPUs 
implement scoreboarding~\cite{HennesseyPatterson}, we were able to 
alleviate this bottleneck by interleaving tree walks. 
% This significantly improved performance of generated 
% code. The fact that exploiting ILP improved performance on GPUs was surprising.

>>>>

\section{Overview of \Treebeard{} Multi-Level Compilation}

@@

\Treebeard{} takes a serialized decision tree ensemble as input (
XGBoost JSON, ONNX etc.) and automatically generates an optimized inference function
that can either target CPUs or GPUs. 
Figure \ref{Fig:CompilerStructure} shows the structure of the \Treebeard{} compiler. 
The inference computation is lowered through three intermediate representations
-- high-level IR (HIR), mid-level IR (MIR) and low-level IR (LIR). The LIR is
finally lowered to LLVM and then JIT'ed to the specified target processor. 
\Treebeard{} is built using the open-source \TreebeardOLD{} infrastructure~\cite{Treebeard}.
% Since the \TreebeardOLD{} infrastructure was originally designed to target CPUs,
% significant extensions were required to support schedule-guided compilation for CPUs and GPUs. 
The \TreebeardOLD{} infrastructure was originally designed to target CPUs.
It lacks a scheduling language and does not support generating code for 
different implementation strategies. We extend \TreebeardOLD{} significantly 
to support schedule-guided compilation for CPUs and GPUs.
The parts \Treebeard{} that are new or significantly different compared to \TreebeardOLD{}
are shown as shaded boxes in Figure \ref{Fig:CompilerStructure}.
% However, it implements several optimizations on the HIR and MIR that 
% can be leveraged across target processors and we find that some these 
% are beneficial for GPUs as well. 
% This reuse of optimizations is possible 
% because the intermediate representations on which these optimizations are performed 
% are abstract and are designed to be target-independent. We briefly review these 
% optimizations below.

@@

The HIR is lowered to MIR as dictated by the \emph{schedule}.
Optimizations like tree-walk unrolling and tree-walk interleaving
are performed on the MIR. 
% These optimizations are beneficial for GPUs as well.
% and the design of \TreebeardOLD{} allows us to reuse the tree-walk 
% unrolling and tree-walk interleaving optimizations on the MIR for GPUs.
% While building \Treebeard{}, we found that
One surprising thing we found while developing \Treebeard{} was that 
we could use ILP to improve performance on GPUs. 
One of the performance bottlenecks in inference code targeted 
to GPUs was that warps spent significant time being stalled. We were able to 
alleviate this bottleneck by interleaving tree walks. 
% This significantly improved performance of generated 
% code. The fact that exploiting ILP improved performance on GPUs was surprising.

### 2024-04-17 docs/SOSP 2024/schedule.tex
Edit overview and schedule sections

\section{Scheduling Language}
\label{sec:schedule}
As established in Section \ref{sec:motivation}, there are several 
configurations and optimizations strategies for decision 
tree inference. The best one to use varies significantly across
models, batch sizes and hardware platforms. Therefore, designing
any one hard-coded strategy is not feasible as this would 
make portable performance impossible. To address this, we 
design a scheduling language for \Treebeard{}. The scheduling 
language provides an abstract way to specify loop structure and 
other optimizations as an input to the compiler. The specified 
\textbf{\emph{schedule}} controls the lowering of model inference to a set of 
loop nests. The configurability provided by the schedule allows us 
to build auto-schedulers and auto-tuners (Section \ref{sec:exploring}).
% Using a scheduling language will also significantly simplify 
% adding support for new hardware as this will likely require
% different locality optimizations and loop transformations.

@@

% \subsection{Language Definition}
The core construct of \Treebeard{}'s scheduling language is an 
\textbf{\emph{index variable}} which abstractly represents a loop. 
Each index variable has a range of values that it can take along 
with a step. 
The language provides directives to manipulate these index 
variables. There are two special index variables -- \op{batch} and
\op{tree} that are used to represent the batch and tree loops and all 
other index variables are derived from these. A schedule derives 
new index variables from these root index variables by applying
directives. \TODO{Should we somehow explain what tree and batch loop are?}

@@

\Treebeard{}'s scheduling language is expressive enough to represent a wide range of
strategies used in existing systems. We show examples of how it can be used to 
represent XGBoost and Tahoe's strategies. 
% Before presenting the examples, we note that \Treebeard{}'s 
% default loop order is [\op{batch}, \op{tree}], i.e, for each row in the input
% batch, go over all trees.

% \subsection{The XBoost Schedule}
XGBoost\cite{XGBoost} implements inference on the CPU by going 
over a fixed number of rows (64 in the previous version)
for every tree and then moving to the next tree. It moves to the 
next set of rows when all trees have been walked for the current 
set of rows. Different sets of rows are processed in parallel.
The following schedule expresses XGBoost's strategy.
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, CHUNK_SIZE)
  reorder(b0, tree, b1)
  parallel(b0)
\end{lstlisting}

@@

In the \emph{direct method}, a single GPU thread walks all trees
for a given input row. The schedule for this strategy is as follows.
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, ROWS_PER_TB)
  reorder(b0, b1, tree)
  gpuDimension(b0, grid.x)
  gpuDimension(b1, block.x)
\end{lstlisting}
Here, \op{ROWS\_PER\_TB} is the number of rows that are processed by a single thread block.

In the \emph{shared data} strategy, a thread block walks all the trees 
for a given row in parallel. 
% If threads walk multiple trees, each thread accumulates partial results. 
Then, a thread block wide reduction is performed to compute 
the prediction. The schedule for this strategy is as follows.
\begin{lstlisting}[style=c++]
reorder(batch, tree)
gpuDimension(batch, grid.x)
gpuDimension(tree, block.x)
cache(batch)
\end{lstlisting}
%   \item \textbf{Shared Forest}: In this strategy, the whole model is loaded into 
%   shared memory and subsequently, a single thread walks all trees for a particular
%   row. The schedule for this strategy is as follows.
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, ROWS_PER_TB)
%   tile(tree, t0, t1, N_TREES)
%   reorder(b0, b1, t0, t1)
%   cache(t0)
%   gpuDimension(b0, grid.x)
%   gpuDimension(b1, block.x)
% \end{lstlisting}
%   Here, we create a placeholder single iteration loop \op{t0} so that we can 
%   specify that all trees are to be cached.
%   \item \textbf{Shared Partial Forest}: In case the model is too large to fit into
%   shared memory, the model is split into chunks and each chunk is loaded into shared
%   memory. Again, as in the previous strategy, one thread walks all trees assigned to
%   a thread block for a row. The schedule for this strategy is as follows.
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, ROWS_PER_TB)

>>>>

\section{Scheduling Language}
\label{sec:schedule}
% As established in Section \ref{sec:motivation}, there are several 
% configurations and optimizations strategies for decision 
% tree inference. The best one to use varies significantly across
% models, batch sizes and hardware platforms. Therefore, designing
% any one hard-coded strategy is not feasible as this would 
% make portable performance impossible. 
We design a scheduling language for \Treebeard{}
to address the problems discussed in Section \ref{sec:motivation}. The scheduling 
language provides an abstract way to specify loop structure and 
other optimizations as an input to the compiler. The specified 
\textbf{\emph{schedule}} controls the lowering of model inference to a set of 
loop nests. The configurability provided by the schedule allows us 
to build auto-schedulers and auto-tuners (Section \ref{sec:exploring}).
% Using a scheduling language will also significantly simplify 
% adding support for new hardware as this will likely require
% different locality optimizations and loop transformations.

@@

\subsection{Language Definition}
The core construct of \Treebeard{}'s scheduling language is an 
\textbf{\emph{index variable}} which abstractly represents a loop. 
Each index variable has a range of values that it can take along 
with a step. 
The language provides directives to manipulate these index 
variables. There are two special index variables -- \op{batch} and
\op{tree} that are used to represent the batch and tree loops and all 
other index variables are derived from these. A schedule derives 
new index variables from these root index variables by applying
directives. \TODO{Should we somehow explain what tree and batch loop are?}

@@

\subsection{Expressiveness of the Scheduling Language}
\Treebeard{}'s scheduling language is expressive enough to represent a wide range of
strategies used in existing systems. We show examples of how it can be used to 
represent XGBoost~\cite{XGBoost} and Tahoe's strategies~\cite{Tahoe}. 
% Before presenting the examples, we note that \Treebeard{}'s 
% default loop order is [\op{batch}, \op{tree}], i.e, for each row in the input
% batch, go over all trees.

XGBoost\cite{XGBoost} implements inference on the CPU by going 
over a fixed number of rows (64 in the previous version)
for every tree and then moving to the next tree. It moves to the 
next set of rows when all trees have been walked for the current 
set of rows. Different sets of rows are processed in parallel.
This is expressed in \Treebeard{}'s scheduling language as:
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, CHUNK_SIZE)
  reorder(b0, tree, b1)
  parallel(b0)
\end{lstlisting}

@@

In the \emph{direct method}~\cite{Tahoe}, a single GPU thread walks all trees
for a given input row. The schedule for this strategy is as follows.
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, ROWS_PER_TB)
  reorder(b0, b1, tree)
  gpuDimension(b0, grid.x)
  gpuDimension(b1, block.x)
\end{lstlisting}
Here, \op{ROWS\_PER\_TB} is the number of rows that are processed by a single thread block.

In the \emph{shared data} strategy~\cite{Tahoe}, a thread block walks all the trees 
for a given row in parallel. 
% If threads walk multiple trees, each thread accumulates partial results. 
Then, a thread block wide reduction is performed to compute 
the prediction. The schedule for this strategy is as follows.
\begin{lstlisting}[style=c++]
reorder(batch, tree)
gpuDimension(batch, grid.x)
gpuDimension(tree, block.x)
cache(batch)
\end{lstlisting}
%   \item \textbf{Shared Forest}: In this strategy, the whole model is loaded into 
%   shared memory and subsequently, a single thread walks all trees for a particular
%   row. The schedule for this strategy is as follows.
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, ROWS_PER_TB)
%   tile(tree, t0, t1, N_TREES)
%   reorder(b0, b1, t0, t1)
%   cache(t0)
%   gpuDimension(b0, grid.x)
%   gpuDimension(b1, block.x)
% \end{lstlisting}
%   Here, we create a placeholder single iteration loop \op{t0} so that we can 
%   specify that all trees are to be cached.
%   \item \textbf{Shared Partial Forest}: In case the model is too large to fit into
%   shared memory, the model is split into chunks and each chunk is loaded into shared
%   memory. Again, as in the previous strategy, one thread walks all trees assigned to
%   a thread block for a row. The schedule for this strategy is as follows.
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, ROWS_PER_TB)
